\documentclass[acmsmall,colorlinks, dvipsnames]{acmart}
%% colorlinks: option for package hyperref
%% dvipsnames: option for package xcolor

\usepackage{mystyle}
\usepackage{mymacros}
\usepackage{booktabs}
\usepackage{xtab}

% Stretch whitespace a little bit to avoid overfull.
\emergencystretch=1em

\setcitestyle{numbers, comma}

%%Este comando eh usado para referenciar código PVS
% \newcommand{\pvscode}[1]{\texttt{#1}} 
\newcommand{\pvscode}[1]{\ensuremath{\mathit{#1}}} 
\newtheorem{axiom}{Axiom}

%\journal{Science of Computer Programming}

%%
%% These commands are for a JOURNAL article.
\acmJournal{TOSEM}

%% The following commands result in compilation errors if given empty arguments.
% \acmVolume{}
% \acmNumber{}
% \acmArticle{}
% \acmMonth{}



%\bibliographystyle{elsarticle-harv}
\bibliographystyle{model2-names}


\begin{document}
 
\selectlanguage{english}

%\begin{frontmatter}



\title{dA Formal Framework of Software Product Line Analyses}

    % Authors
    
%Thiago
%Leopoldo
%Vander
%Sven
%Maxime
%Rohit



\author{Thiago Castro}
\affiliation{%
  \institution{Systems Development Center - Brazilian Army}
  \streetaddress{QG do Exército - Bloco G - 2º Andar, 70630-901, Setor Militar Urbano}
  \city{Brasilia}
  \country{Brazil}}
\email{thiago.mael@gmail.com}

\author{Leopoldo Teixeira}
\affiliation{%
  \institution{Federal University of Pernambuco}
  \streetaddress{Av. Jornalista Aníbal Fernandes, s/n - Cidade Universitária (Campus Recife), 50740-560}
  \city{Recife}
  \country{Brazil}}
\email{lmt@cin.ufpe.br}

\author{Vander Alves}
\affiliation{%
  \institution{University of Brasília}
  \streetaddress{Campus Universitário Darcy Ribeiro - Edifício CIC/EST, 70910-900}
  \city{Brasilia}
  \country{Brazil}}
\email{valves@unb.br}

\author{Sven Apel}
\affiliation{%
  \institution{Saarland University, Saarland Informatics Campus}
  \streetaddress{Campus E1 1, 66123} 
%  \streetaddress{Saarland Informatics Campus, Saarland University, Campus E1 1, 66123}
  \city{Saarbruecken}
  \country{Germany}}
\email{apel@cs.uni-saarland.de}

\author{Maxime Cordy}
\affiliation{%
  \institution{University of Luxembourg}
  \streetaddress{6, rue Richard Coudenhove-Kalergi L-1359}
  \city{Luxembourg}
  \country{Luxembourg}}
\email{maxime.cordy@uni.lu}

\author{Rohit Gheyi}
\affiliation{%
  \institution{Federal University of Campina Grande}
  \streetaddress{Av. Aprigio Veloso, 882, Bloco CN, Bairro Universitário
58.429-900}
  \city{Campina Grande}
  \country{Brazil}}
\email{rohit@dsc.ufcg.edu.br}




    


%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}

A number of approaches lift analyses such as type checking, model
checking, and theorem proving to product-line analysis. These
approaches share features that suggest an unexplored potential for
reuse of key analysis steps and properties,
implementation, and verification efforts. Despite the availability of
taxonomies synthesizing such approaches, there still remains the
underlying problem of not being able to describe precisely and
uniformly product-line analyses and their properties. We
propose a formal framework of product-line analysis. 
The framework models analyses in a compositional
manner, providing an overall understanding of the space of
family-based, feature-based, and product-based analysis strategies. 
It defines precisely how the different types of product-line analyses compose and
inter-relate. To ensure soundness,  we formalize the framework, 
providing a precise specification of key concepts and properties of
the individual analyses as well as proofs of key soundness results. 
The formalization provides unambiguous definitions of domain terminology and assumptions as well as solid evidence of key properties based on rigorous formal proofs. 
To qualitatively assess the generality of the framework, we
 discuss to what extent it describes five
representative product-line analyses on the following properties:
safety, performance, data-flow facts, security, and functional program
properties. 



% Problem
%There is a number of approaches that lift analyses such as type checking, model checking, and theorem proving to \emph{product-line analysis}. Although different in purpose, these approaches share features that suggest there is a yet unexplored potential for reuse of specification of key analysis steps and properties, implementation, and verification efforts.
% Relevance
%Despite the availability of taxonomies synthesizing such approaches, 
%there still remains the underlying problem of not being able to describe precisely and uniformly product-line analyses and their properties. Indeed, more refinement is necessary to formalize core concepts, explore similarities in analysis steps and key properties, and suitably manage variability among these approaches.
% Solution 
%To address this underlying problem, we propose a framework of product-line analysis and provide guidance on its customization. The framework presents  analyses in a compositional manner at a conceptual level,  
%providing an overall understanding of the structure of the space of family-based, feature-based, and product-based analyses, showing how the different types of product-line analyses compose and inter-relate.  
% Evaluation
%To ensure soundness of the framework, we formalize it using the PVS proof assistant, providing a precise specification of key concepts and properties of the individual analyses as well as proofs of key soundness results. 
%To qualitatively assess the generality of the framework, we retrospectively discuss to what extent it can describe five representative product-line analyses on the following properties: safety, performance, data-flow facts, security, and functional program properties. 
%Lastly, to guide the use of the framework, we define a product line of product-line analyses techniques 
%($PL^{2}Ana$), which exposes the configurability of the framework and systematizes its customization. Overall, the framework provides a more comprehensive and deeper understanding of underlying 
%principles used in these analyses, which we envision could help other researchers to lift existing 
%single-product analysis techniques to yet under-explored variability-aware approaches. 

\end{abstract}

%%
%% The code below is generated by the tool at http://dl.acm.org/ccs.cfm.
%%
\begin{CCSXML}
<ccs2012>
   <concept>
       <concept_id>10011007.10011074.10011092.10011096.10011097</concept_id>
       <concept_desc>Software and its engineering~Software product lines</concept_desc>
       <concept_significance>500</concept_significance>
       </concept>
   <concept>
       <concept_id>10011007.10011074.10011099.10011692</concept_id>
       <concept_desc>Software and its engineering~Formal software verification</concept_desc>
       <concept_significance>500</concept_significance>
       </concept>
 </ccs2012>
\end{CCSXML}

\ccsdesc[500]{Software and its engineering~Software product lines}
\ccsdesc[500]{Software and its engineering~Formal software verification}

\keywords{Software Product Lines, Product-line Analysis, Framework, Formalization}

%%
%% This command processes the author and affiliation and title
%% information and builds the first part of the formatted document.
\maketitle
    

%    \input{content/abstract}

%    \begin{keyword}
%        Software Product Lines\sep Product-line Analysis \sep Verification
%    \end{keyword}

%\end{frontmatter}

\input{content/introduction}
\input{content/background}
\input{content/strategies}
\input{content/framework}
%\input{content/pl2ana}
\input{content/threats}
\input{content/relatedWork}
\input{content/conclusion}

\section*{Acknowledgements}

We would like to thank the following people for fruitful discussions and suggestions on how to improve this work: Tobias Sena, Andreas Stahlbauer, Christoph Seidl, Malte Lochau, Matthias Kowal, Ina Schaefer, Thomas Th{\"u}m, and Pierre-Yves Schobbens. 
Vander Alves was partially supported by CNPq (grant 310757/2018-5), FAPDF (grant SEI 00193-00000926/2019-67), and the Alexander von Humboldt Foundation.
Leopoldo Teixeira was partially supported by CNPq (grant 409335/2016-9) and FACEPE (APQ-0570-1.03/14), as well as INES 2.0,\footnote{\url{http://www.ines.org.br}} FACEPE grants PRONEX APQ-0388-1.03/14 and APQ-0399-1.03/17, and CNPq grant 465614/2014-0.
Sven Apel was partially supported by the German Research Foundation (DFG) within the Heisenberg Programme (AP 206/6).

%\section*{\refname}
%\bibliography{references}
\bibliography{references,mcr}
%qualifying

%\newpage
%\begin{appendices}
%
%\renewcommand\thefigure{\thesection.\arabic{figure}}
%\setcounter{figure}{0}
%
%\section{Probabilistic Models}
%\label{app:probabilistic-models}
%\input{content/appendix/models}
%
%\end{appendices}

\end{document}
